---
title: Gene Ontology Enrichment/Overrepresentation Analysis on Pulmonary TB Infected
  Tissues in Mice
project for: Personal entertainment and study
author/copyright(c): Wani2Y
output:
  html_document:
    code_folding: show
    toc: yes
    number_section: yes
    theme: darkly
    highlight: zenburn
    css: style.css
last modified: 07/08/2024
first created: 01/02/2025
---

```{r doc setup, eval=TRUE, include=FALSE}
knitr::opts_chunk$set(
	message = FALSE,
	warning = FALSE,
	cache = TRUE,
	cache.lazy = FALSE,
	collapse = TRUE,
	out.width = "100%",
	results = "hold",
	tidy.opts=list(width.cutoff = 112),
	tidy = TRUE
)
```

# Disclaimer
|               This notebook is designed to explore the molecular data published at DOI: 10.1128/iai.00263-23; This notebook is free to use; you can redistribute it and/or modify it under the terms of the GNU General Public License,version 3, as published by the Free Software Foundation. This notebook is distributed in the hope that it will serve educational and entertainment purposes, but without any warranty; without even the implied warranty of merchantability or fitness for a particular purpose. Use of this code may be subjected to licensing agreement and copyright requirement of the Journal, Infection and Immunity.
|               For more details of the license, please refer to: www.gnu.org/licenses/gpl-3.0.en.html.

# Environment setup
```{r install the required packages}
# uncomment the following lines of this chunk if you have not installed all required packages. Modify the code if you already have some of the packages.
#if (!requireNamespace("BiocManager", quietly = TRUE))
#  install.packages("BiocManager")

#BiocManager::install(c("stats", "affy", "limma", "annotate", "oligo", "Biobase", "R.utils", "dplyr", "tidyverse", "clusterProfiler", "org.Mm.eg.db", "ggplot2", "enrichplot", "reshape2", "plotly", "GOSemSim"))
```

# Data source
|               Article citation: Koyuncu D, Tavolara T, Gatti DM, Gower AC et al. B cells in perivascular and peribronchiolar granuloma-associated lymphoid tissue and B-cell signatures identify asymptomatic Mycobacterium tuberculosis lung infection in Diversity Outbred mice. Infect Immun 2024 Jul 11;92(7):e0026323. PMID: 38899881.
|               Raw expression microarray data is imported from Gene Expression Omnibus (GEO): https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE266564. Gene Ontology (GO) data is obtained from the associated platform, namely GPL16570, and the R package, org.Mm.eg.db.

# Context
|               The source article experimented on lung samples of Diversity Outbred mice infected by $\textit{Mycobacterium tuberculosis}$ (TB) over maximally 60 days. Three states of infections were identified into groups, namely not infected, asymptomatic controller, chronic controller, and progressor groups with definitions as follows: (1) uninfected mice were not experimentally infected by TB; (2) asymptomatic mice were experimentally infected but lacked diagnosible symptom, which were euthanised at predetermined date at of before 60 days post-infection; (3) chronically infected mice with symptoms of TB, which morbidly succumbed to TB and were euthanized after 60 days post-infection; (4) progressively infected mice with with symptoms of TB, which morbidly succumbed to TB and were euthanized before 60 days post-infection.
|               B-cell pathways and CD20+ B cells were found to be enrich/overrepresented in asymptomatic controller group. This report aims to perform an enrichment/overrepresentation analysis to see whether the same conclusion can be reached. In addition, genes related to the following Gene Ontology are found to be enriched/overrepresented in chronic controller and progressor groups: (1) Antigen receptor-mediated, (2) signaling pathway, (3) B-cell receptor signaling pathway, (4) T-cell activation, and (5) Regulation of lymphocyte activation.
|               From the published data on GEO, the microarray data is constituted of the superseries GSE 266564, which is subdivided into two subseries: GSE 266562 and 266563. Here, we treat the entire 134 samples in the superseries as a single dataset for analysis.

# Potential issues
|               (1) Comparison method: 105 overrepresented genes are not recovered by pairwise comparison. Instead, asymptomatic and the combination of chronic controller, progressor, and non-infected mice are compared.
|               (2) GEO data: A subset of microarray data (i.e. GSE179417) is said to be published elsewhere, which likely refers to the dataset under a different article. The complete microarray data is not linked in the article. The article name locates GSE266564 as the microarray data for this article. 
|               (3) GEO annotation platform:  GSE266564uses platform GPL32068 for annotating the microarray data. GPL32068 is described as identical to GPL16570 but the table dimension attributes and number of rows are not the same. 
|               (4) R script: The R script to perform overrepresentation analysis is not provided in the article (i.e. no file in Supplementary Information or GitHub link).
|               (5) DEG computation: The choice of Mann-Whitney U-test over Welch's t-test and Bayesian moderation is seemly not specified in the article, though we may presume the choice is based on the lack of assumption on normality of the underlying data.
|               (6) Enrichr: Enrichr does not preserve all parent-child relationship among their hierarchy, and a cut-off level is implemented. By comparison, enrichGO does preserve all hierarchy. The two implementations have been found to recovered different number of genes in overrepresentation analysis ([Yu Lab](https://mp.weixin.qq.com/s/6lSsg2WMEK2btwve-9C2rA)).

# Study Objective
|               (1) Perform Exploratory Data Analysis (EDA) on the source CEL files, to test whether the choice of Mann-Whitney U-test over Bayesian moderation is valid based on two criteria: Normality and outliers. 
|               (2) Perform an overrepresentation analysis on Biological Process using [GSE266564](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE266564) and the lung portion of [GPL16570](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GPL16570) related to lung tissues, which is an exploration to recreate the result from the article based on what may be considered as complete and equivalent data.

# Microarry data procurement and cleaning
```{r load in the necessary library, results='hide'}
library(stats)
library(affy)
library(limma)
library(effsize)
library(preprocessCore)
library(Biobase)
library(R.utils)
library(dplyr)
library(stringr)
library(tidyverse)
library(clusterProfiler)
library(org.Mm.eg.db)
library(GOSemSim)
library(ggplot2)
library(enrichplot)
library(reshape2)
library(plotly)
```

## Downloading microarray data from GEO
```{r dowmload the microarray data from GEO, results='hide'}
# <NOTE first run of this chunk may cause error, run it again without modification solve the issue. Debugging required>
# try working on one or two small sets of samples first if you want to try out this analysis.
# mice were infected with different dosage of M. tuberculosis, but we probably do not need to differentiate them for our purpose here.
# define a dataframe as sample reference for the subsequent annotated dataframe
cell_sample <- data.frame(cbind(
  Ex_group = c(rep("uninfected", 5),
               rep("asymptomatic_controller", 14), 
               rep("chronic_controller", 17), #on GEO, it is labelled as "Controller".
               rep("progressor", 10), # on GEO, it is labelled as "Supersusceptible Progressor".
               rep("uninfected", 20),
               rep("asymptomatic_controller", 39),
               rep("progressor", 29)
               ), 
  
  cell_series = c("GSM8250425", "GSM8250426", "GSM8250427", "GSM8250428", "GSM8250429", # uninfected_1
                  "GSM8250430", "GSM8250431", "GSM8250432", "GSM8250433", "GSM8250434", # asymptomatic controller_1
                  "GSM8250435", "GSM8250436", "GSM8250437", "GSM8250438", "GSM8250439", 
                  "GSM8250440", "GSM8250441", "GSM8250442", "GSM8250443", 
                  "GSM8250444", "GSM8250445", "GSM8250446", "GSM8250447", "GSM8250448", # chronic controller_1
                  "GSM8250449", "GSM8250450", "GSM8250451", "GSM8250452", "GSM8250453", 
                  "GSM8250454", "GSM8250455", "GSM8250456", "GSM8250457", "GSM8250458", 
                  "GSM8250459", "GSM8250460", 
                  "GSM8250461", "GSM8250462", "GSM8250463", "GSM8250464", "GSM8250465", # progressor_1
                  "GSM8250466", "GSM8250467", "GSM8250468", "GSM8250469", "GSM8250470", 
                  "GSM8250471", "GSM8250472", "GSM8250473", "GSM8250474", "GSM8250475", # uninfected_2
                  "GSM8250476", "GSM8250477", "GSM8250478", "GSM8250479", "GSM8250480", 
                  "GSM8250481", "GSM8250482", "GSM8250483", "GSM8250484", "GSM8250485",
                  "GSM8250486", "GSM8250487", "GSM8250488", "GSM8250489", "GSM8250490", 
                  "GSM8250491", "GSM8250492", "GSM8250493", "GSM8250494", "GSM8250495", # asymptomatic controller_2
                  "GSM8250496", "GSM8250497", "GSM8250498", "GSM8250499", "GSM8250500",
                  "GSM8250501", "GSM8250502", "GSM8250503", "GSM8250504", "GSM8250505",
                  "GSM8250506", "GSM8250507", "GSM8250508", "GSM8250509", "GSM8250510",
                  "GSM8250511", "GSM8250512", "GSM8250513", "GSM8250514", "GSM8250515",
                  "GSM8250516", "GSM8250517", "GSM8250518", "GSM8250519", "GSM8250520",
                  "GSM8250521", "GSM8250522", "GSM8250523", "GSM8250524", "GSM8250525",
                  "GSM8250526", "GSM8250527", "GSM8250528", "GSM8250529",
                  "GSM8250530", "GSM8250531", "GSM8250532", "GSM8250533", "GSM8250534", # progressor_2
                  "GSM8250535", "GSM8250536", "GSM8250537", "GSM8250538", "GSM8250539",
                  "GSM8250540", "GSM8250541", "GSM8250542", "GSM8250543", "GSM8250544",
                  "GSM8250545", "GSM8250546", "GSM8250547", "GSM8250548", "GSM8250549",
                  "GSM8250550", "GSM8250551", "GSM8250552", "GSM8250553", "GSM8250554",
                  "GSM8250555", "GSM8250556", "GSM8250557", "GSM8250558"
                  ), 
  
  file_tag = c("5F161209", "5F161209", "5F161209", "5F161209", "5F161209", # uninfected_1
               "5F161209", "5F161209", "5F161209", "5F161209", "5F161209",  # asymptomatic controller_1
               "5F161209", "5F161209", "5F161209", "5F191212", "5F161209", 
               "5F161209", "5F161209", "5F161209", "5F191212",
               "5F161209", "5F161209", "5F161209", "5F191212", "5F161209", # chronic controller_1
               "5F161209", "5F161209", "5F191212", "5F161209", "5F161209", 
               "5F191212", "5F161209", "5F191212", "5F191212", "5F191212", 
               "5F191212", "5F191212", 
               "5F161209", "5F161209", "5F161209", "5F161209", "5F161209", # progressor_1
               "5F161209", "5F161209", "5F161209", "5F161209", "5F161209", 
               "5F161209", "5F161209", "5F191212", "5F170526", "5F170526", # uninfected_2
               "5F161209", "5F191212", "5F170526", "5F161209", "5F191212", 
               "5F170526", "5F161209", "5F191212", "5F170526", "5F191212", 
               "5F170526", "5F191212", "5F170526", "5F191212", "5F170526", 
               "5F170526", "5F170526", "5F170526", "5F170526", "5F170526", # asymptomatic controller_2
               "5F191212", "5F170526", "5F170526", "5F191212", "5F191212",
               "5F191212", "5F191212", "5F170526", "5F191212", "5F170526",
               
               "5F170526", "5F191212", "5F170526", "5F191212", "5F170526",
               "5F191212", "5F170526", "5F170526", "5F170526", "5F191212",
               "5F170526", "5F191212", "5F170526", "5F191212", "5F170526",
               "5F170526", "5F170526", "5F170526", "5F170526", "5F191212",
               "5F191212", "5F170526", "5F191212", "5F191212",
               "5F170526", "5F170526", "5F191212", "5F170526", "5F191212", # progressor_2
               "5F191212", "5F191212", "5F170526", "5F191212", "5F191212",
               "5F191212", "5F170526", "5F191212", "5F191212", "5F170526",
               "5F170526", "5F170526", "5F170526", "5F191212", "5F170526",
               "5F170526", "5F170526", "5F170526", "5F170526", "5F170526",
               "5F191212", "5F191212", "5F191212", "5F170526"
             ), 
  
  grp_tag = c("E1", "E2", "E3", "E4", "E5", # uninfected_1
              "A1", "A2", "A3", "A4", "A6",  # asymptomatic controller_1
              "A7", "B1", "B5", "A1", "B7", 
              "C3", "C7", "C8", "B2",
              "A5", "B3", "B4", "A2", "C1", # chronic controller_1
              "C4", "C5", "A3", "D1", "D3", 
              "A5", "D5", "A6", "A7", "A8", 
              "B1", "B3", 
              "A8", "B2", "B6", "B8", "C2", # progressor_1
              "C6", "D2", "D6", "D7", "D8", 
              "E6", "E7", "F2", "F1", "F2", # uninfected_2
              "E8", "F3", "F3", "F1", "F4", 
              "F4", "F2", "F5", "F5", "F6", 
              "F6", "F7", "F7", "F8", "F8", 
              "A2", "A4", "A7", "B1", "B2", # asymptomatic controller_2
              "C3", "B3", "B5", "C6", "C7",
              "C8", "D2", "C2", "D4", "C3",
              "C4", "D5", "C7", "D6", "C8",
              "D7", "D1", "D2", "D4", "D8",
              "D6", "E1", "D7", "E2", "E2",
              "E3", "E4", "E5", "E6", "E5",
              "E6", "E7", "E7", "F1",
              "A1", "A3", "B4", "A5", "B5", # progressor_2
              "B6", "B7", "A6", "B8", "C1",
              "C2", "A8", "C5", "D1", "B6",
              "B7", "B8", "C1", "D3", "C5",
              "C6", "D3", "D5", "D8", "E1",
              "E3", "E4", "E8", "E8"
              ), 
  
  file_name = c("F249", "F250", "F251", "F252", "F253", # uninfected_1
                "F194", "F195", "F196", "F197", "F199", # asymptomatic controller_1
                "F200", "F202", "F206", "F206", "F208", 
                "F213", "F217", "F218", "F236", 
                "F198", "F204", "F205", "F209", "F211", # chronic controller_1
                "F214", "F215", "F215", "F220", "F222", 
                "F225", "F227", "F229", "F232", "F233", 
                "F234", "F241", 
                "F201", "F203", "F207", "F210", "F212", # progressor_1
                "F216", "F221", "F231", "F239", "F240", 
                "F474", "F475", "F475", "F475", "F476", # uninfected_2
                "F477", "F477", "F477", "F478", "F478", 
                "F478", "F479", "F479", "F479", "F480", 
                "F480", "F481", "F481", "F482", "F482", 
                "F260", "F265", "F288", "F296", "F300", # asymptomatic controller_2
                "F303", "F305", "F311", "F318", "F329", 
                "F336", "F359", "F359", "F367", "F368",
                "F372", "F378", "F384", "F386", "F386",
                "F389", "F394", "F398", "F409", "F415",
                "F417", "F418", "F418", "F422", "F424",
                "F426", "F429", "F440", "F443", "F451",
                "F452", "F455", "F459", "F467",
                "F256", "F261", "F267", "F272", "F274", # progressor_2
                "F279", "F281", "F285", "F286", "F295",
                "F299", "F299", "F316", "F340", "F344",
                "F350", "F351", "F355", "F365", "F375",
                "F383", "F407", "F411", "F419", "F421",
                "F434", "F437", "F465", "F471"
                ))) 

# set row names for subsequent indexing.
rownames(cell_sample) <- c("GSM8250425.CEL", "GSM8250426.CEL", "GSM8250427.CEL", "GSM8250428.CEL", "GSM8250429.CEL", # uninfected_1
                           "GSM8250430.CEL", "GSM8250431.CEL", "GSM8250432.CEL", "GSM8250433.CEL", "GSM8250434.CEL", # asymptomatic controller_1
                           "GSM8250435.CEL", "GSM8250436.CEL", "GSM8250437.CEL", "GSM8250438.CEL", "GSM8250439.CEL", 
                           "GSM8250440.CEL", "GSM8250441.CEL", "GSM8250442.CEL", "GSM8250443.CEL", 
                           "GSM8250444.CEL", "GSM8250445.CEL", "GSM8250446.CEL", "GSM8250447.CEL", "GSM8250448.CEL", # chronic controller_1
                           "GSM8250449.CEL", "GSM8250450.CEL", "GSM8250451.CEL", "GSM8250452.CEL", "GSM8250453.CEL", 
                           "GSM8250454.CEL", "GSM8250455.CEL", "GSM8250456.CEL", "GSM8250457.CEL", "GSM8250458.CEL", 
                           "GSM8250459.CEL", "GSM8250460.CEL", 
                           "GSM8250461.CEL", "GSM8250462.CEL", "GSM8250463.CEL", "GSM8250464.CEL", "GSM8250465.CEL", # progressor_1
                           "GSM8250466.CEL", "GSM8250467.CEL", "GSM8250468.CEL", "GSM8250469.CEL", "GSM8250470.CEL", 
                           "GSM8250471.CEL", "GSM8250472.CEL", "GSM8250473.CEL", "GSM8250474.CEL", "GSM8250475.CEL", # uninfected_2
                           "GSM8250476.CEL", "GSM8250477.CEL", "GSM8250478.CEL", "GSM8250479.CEL", "GSM8250480.CEL", 
                           "GSM8250481.CEL", "GSM8250482.CEL", "GSM8250483.CEL", "GSM8250484.CEL", "GSM8250485.CEL", 
                           "GSM8250486.CEL", "GSM8250487.CEL", "GSM8250488.CEL", "GSM8250489.CEL", "GSM8250490.CEL", 
                           "GSM8250491.CEL", "GSM8250492.CEL", "GSM8250493.CEL", "GSM8250494.CEL", "GSM8250495.CEL", # asymptomatic controller_2
                           "GSM8250496.CEL", "GSM8250497.CEL", "GSM8250498.CEL", "GSM8250499.CEL", "GSM8250500.CEL",
                           "GSM8250501.CEL", "GSM8250502.CEL", "GSM8250503.CEL", "GSM8250504.CEL", "GSM8250505.CEL",
                           "GSM8250506.CEL", "GSM8250507.CEL", "GSM8250508.CEL", "GSM8250509.CEL", "GSM8250510.CEL",
                           "GSM8250511.CEL", "GSM8250512.CEL", "GSM8250513.CEL", "GSM8250514.CEL", "GSM8250515.CEL",
                           "GSM8250516.CEL", "GSM8250517.CEL", "GSM8250518.CEL", "GSM8250519.CEL", "GSM8250520.CEL",
                           "GSM8250521.CEL", "GSM8250522.CEL", "GSM8250523.CEL", "GSM8250524.CEL", "GSM8250525.CEL",
                           "GSM8250526.CEL", "GSM8250527.CEL", "GSM8250528.CEL", "GSM8250529.CEL",
                           "GSM8250530.CEL", "GSM8250531.CEL", "GSM8250532.CEL", "GSM8250533.CEL", "GSM8250534.CEL", # progressor_2
                           "GSM8250535.CEL", "GSM8250536.CEL", "GSM8250537.CEL", "GSM8250538.CEL", "GSM8250539.CEL",
                           "GSM8250540.CEL", "GSM8250541.CEL", "GSM8250542.CEL", "GSM8250543.CEL", "GSM8250544.CEL",
                           "GSM8250545.CEL", "GSM8250546.CEL", "GSM8250547.CEL", "GSM8250548.CEL", "GSM8250549.CEL",
                           "GSM8250550.CEL", "GSM8250551.CEL", "GSM8250552.CEL", "GSM8250553.CEL", "GSM8250554.CEL",
                           "GSM8250555.CEL", "GSM8250556.CEL", "GSM8250557.CEL", "GSM8250558.CEL"
                           ) 

# download the gene data from GEO
geo <- "https://www.ncbi.nlm.nih.gov/geo/download/"

# iterate through all download links of each sample.
for (ID in cell_sample$cell_series){
  # paste a string for the url
  url <- paste0(geo, "?acc=", 
                ID, 
                "&format=file&file=", 
                ID, 
                "%5FGB%",
                cell_sample[ID, 'file_tag'],
                "%5F", 
                cell_sample[ID, 'grp_tag'], 
                "%5", 
                cell_sample[ID, 'file_name'], 
                "%2ECEL%2Egz")
  # create a file pointer
  fp <- paste0("gene_data/", 
               ID, 
               ".CEL.gz") 
  # check if files already exist
  if (!file.exists(fp)){
    download.file(url, fp)
    Sys.sleep(10) # wait for 5 - 10 seconds between downloads as recommended by GEO. Be nice to the server!
  }
}
```

## Read in Raw Data
```{r load microarray data, results='hide'}
# using ReadAffy() from the affy package would be more concise but it is not applicable for binary cell format of the source article...
# list and unzip .CEL.gz files
cell_fp <- "gene_data/"
cell_name <- list.files(cell_fp, pattern = "\\.CEL\\.gz$", full.names = TRUE)
cell_file <- gsub("\\.gz$", "", cell_name)
for (i in seq_along(cell_name)) {
   if (!file.exists(cell_file[i])) {
    gunzip(cell_name[i], destname = cell_file[i])
  }
}

# assuming you may have loaded packages other than what are included in this document, you may want to specify the oligo package for the read.celfile() function to avoid potential confusions with other packages like affyio.
file_list <- list.celfiles(cell_fp, full.names = TRUE)

cell <- oligo::read.celfiles(cell_file)

# initial look at raw data
raw_exp_matrix <- exprs(cell)  
corr_matrix <- cor(raw_exp_matrix, method = "pearson")

# make matrix into long format
corr_data <- as.data.frame(as.table(corr_matrix))
colnames(corr_data) <- c("Sample1", "Sample2", "Correlation")

p_heat <- ggplot(corr_data, aes(x = Sample1, y = Sample2, fill = Correlation)) +
  geom_tile() +
  scale_fill_gradient2(low = "#041E42", high = "#FF4C00", mid = "white", midpoint = 0.5) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  labs(
    title = "Correlation Heatmap",
    x = "Samples",
    y = "Samples",
    fill = "Correlation"
  )
p_heat
```
|               The raw data is highly correlated, which may suppress subsequent analyses. I have a bad feeling about this...

```{r normalise raw data}
# preprocess raw data for subsequent analyses
# normalise with Robust Multi-Array Average expression(rma).
cell_norm <- rma(cell)

# modify cell_sample into an annotated dataframe.
pheno_data <- AnnotatedDataFrame(data = cell_sample)

# wrangle the data into an expression data.
exp_set <- ExpressionSet(assayData = exprs(cell_norm), 
                      phenoData = pheno_data)
```

# Exploratory Data Analysis (EDA)
|               EDA in this section focus on normality and outlier, which describes the data structure of the source data. 
## Examination on Normality
```{r violin and QQ plots to visually check on normality}
# source data contains too many rows, random down sampling to help visualisation
set.seed(255)
sampled_probe <- sample(1:nrow(exprs(exp_set)), 10)

# create a violin plot for the sample
vio_sample <- exprs(exp_set)[sampled_probe, ]

# transform the wide data into long for violin plot
exp_long <- melt(vio_sample)
colnames(exp_long) <- c("Probe", "Sample", "Expression")

p_vio_sample <- ggplot(exp_long, aes(x = factor(Probe), y = Expression)) +
  geom_violin(fill = "#ADD8E6", color = "#041E42", alpha = 0.7) +
  labs(title = "Violin Plot of Expression Levels Across Randomly Sampled Probes",
       x = "Probes (10 Random Sample)", y = "Expression Levels") +
  theme(axis.text.x = element_blank(),  
        axis.ticks.x = element_blank())
p_vio_sample

# create a qq plot for the sample
qq_sample <- exprs(exp_set)[sampled_probe, ]

qq_data <- do.call(rbind, lapply(1:nrow(qq_sample), function(i) { 
  data.frame(Sample = sort(qq_sample[i, ]),
             Theoretical = qnorm(ppoints(length(qq_sample[i, ]))),
             Probe = paste("Probe", i))
}))

# add a red dotted line to represent the theoretical normal distribution give the data
p_qq_sample <- ggplot(qq_data, aes(x = Theoretical, y = Sample)) +
  geom_point(size = 0.7, color = "#041E42") +
  geom_abline(slope = 1, intercept = 0, color = "#FF4C00", linetype = "dashed") +
  facet_wrap(~ Probe, scales = "free", ncol = 3) +
  labs(title = "Faceted QQ-Plots for Multiple Probes",
       x = "Theoretical Quantiles", y = "Sample Quantiles") +
  theme_minimal()
p_qq_sample
```
```{r use Shapiro-Wilk Test for Normality}
# apply the Shapiro-Wilk test for each probe
shapiro_results <- apply(exprs(exp_set), 1, function(x) shapiro.test(x)$p.value)

# count the number of probes that are not considered, using 0.05 as a cut-off 
non_normal_probe_count <- paste("proportion of non-normal probes:", 
                                length(shapiro_results[shapiro_results < 0.05])/length(shapiro_results))
print(non_normal_probe_count)
```
|               The 10 samples appear somewhat normal on violin plot, but the QQ-plot shows decent skewness in the data away from a theretical normal distribution. Evaluating all probes in a computational method (i.e. Shapiro-Wilk test) shows over half of the probes are not normally distributed.

## Examination on Outliers
```{r box plot}
# visualise the expression data with a boxplot
# put intensity data into a dataframe
box_data <- as.data.frame(exprs(exp_set))

# make the data long
box_data$probe <- rownames(box_data) 
box_data_long <- melt(box_data, 
                      id.vars = "probe", 
                      variable.name = "Sample", 
                      value.name = "Intensity")

# split the samples into subsets for plotting.
n <- 10  
data_chunk <- split(unique(box_data_long$Sample), 
                       ceiling(seq_along(unique(box_data_long$Sample)) / n))

# plotly container
plot_list <- list()

# make plotly objects for all samples by data_chunk
for (i in seq_along(data_chunk)) {
  subset_data <- box_data_long[box_data_long$Sample %in% data_chunk[[i]], ]
  
  p_box <- ggplot(subset_data, aes(x = Sample,
                                   y = Intensity)) +
    geom_boxplot(notch = TRUE) +
    theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
    labs(title = paste("Expression Values by Sample - Set", i),
         x = "Sample", 
         y = "Expression Intensity")
  
  plot_list[[i]] <- ggplotly(p_box)
}

# show them in boxplot and see if any major issues from normalisation
for (p in plot_list) {
  print(p)
} 
```
```{r counting outliers based on quantiles}
outlier_count <- apply(exprs(exp_set), 1, function(x) { # r use 1 and 2 instead of 0 and 1
  Q1 <- quantile(x, 0.25)
  Q3 <- quantile(x, 0.75)
  IQR <- Q3 - Q1
  sum(x < (Q1 - 1.5 * IQR) | x > (Q3 + 1.5 * IQR)) # using a 1.5 IQR to balance sensitivity and robusteness 
})

outlier_percent <- paste("percentage of outliers across probes:", sum(outlier_count)/(nrow(exprs(exp_set))*(ncol(exprs(exp_set)))))
print(outlier_percent)
```
|               We can see a decent number of outliers from each box plot, but the overall percentage of outliers is low.

## Section Conclusion
|               Given the EDA, we have non-normal data with rare outliers. This confirms that the choice of Mann-Whitney U-test in the source article over Welch's t-test or Bayesian moderation is justified.

# Differential Gene Expression (DEG)
|               DEGs in this section look at the expression data from the microarray, and determines which probes from the microarray, if any, are expressed at significantly higher or lower levels. We use the log-fold changes as a measure. Note that for small dataset, or dataset with many outliers empirical Bayes moderation may perform better. 
## Setting up Pairwise Comparison of any Two Groups
```{r define a function}
two_grp_degs <- function(group_1, group_2){
  # Convert input to character.
  grp_1 <- as.character(group_1)
  grp_2 <- as.character(group_2)

  # get the phenodata.
  pheno <- pData(exp_set)[pData(exp_set)$Ex_group %in% c(grp_1, grp_2), ]
  exp_subset <- exprs(exp_set)[, rownames(pheno)]

  probe_results <- data.frame(Probe = rownames(exp_subset), 
                               p.value = NA, 
                               W = NA)

  # Mann-Whitney U-test and cliff delta for each probe
  for (i in 1:nrow(exp_subset)) {
    probe_values <- exp_subset[i, ]
    group_1_values <- probe_values[pheno$Ex_group == grp_1]
    group_2_values <- probe_values[pheno$Ex_group == grp_2]
    
    # Mann-Whitney
    test_result <- wilcox.test(group_1_values, group_2_values, exact = FALSE)
    probe_results$p.value[i] <- test_result$p.value
    probe_results$W[i] <- test_result$statistic
    
    # Cliff delta
    cliff_delta_result <- cliff.delta(group_1_values, group_2_values)
    probe_results$CliffDelta[i] <- cliff_delta_result$estimate
  }

  # sort probes by p-value
  probe_results <- probe_results[order(probe_results$p.value), ]

  # Make a volcano plot to see how many probes are differentially expressed and whether they are up or down regulated
  p_title <- paste0(toupper(substring(grp_1, 1, 1)), tolower(substring(grp_1, 2)), "  VS  ",
                    toupper(substring(grp_2, 1, 1)), tolower(substring(grp_2, 2)))
  # we do not have significant number of outliers, mean instead of median is used here
  probe_results$logFC <- apply(exp_subset, 1, function(x) mean(x[pheno$Ex_group == grp_2]) - mean(x[pheno$Ex_group == grp_1]))
  probe_results$adj.P.Val <- p.adjust(probe_results$p.value, method = "BH")
  x <- probe_results$logFC
  y <- -log2(probe_results$adj.P.Val)
  p <- ggplot(probe_results, aes(x, y)) +
    geom_point() +
    labs(
      x = "LogFC",
      y = "Negative log2 adjusted p-value",
      title =  p_title)+
    geom_hline(yintercept = -log2(0.05), linetype = "dashed", color = "#FF4C00") +
    geom_hline(yintercept = -log2(0.01), linetype = "dashed", color = "#ADD8E6") +
    theme_minimal()

  # Put results together in a list
  result <- list(probe_results, p)
  names(result) <- c("deg_results", "plot")

  # Return the data as list for subsequent cbind
  return(result)
}
```

## Uninfected vs Asymtomatic Controller
```{r perform deg between uninfected and Asymtomatic controller, warning=FALSE}
# note that this section contains many probes that are fully disjoint between the chosen two groups, the warning messages are not printed to avoid clutter. The total disjoint is likely related to the uneven sample size in the source data, which may be accommodated by downsamping if needed.
g <- c("uninfected", "asymptomatic_controller")
deg_ui_ac <- two_grp_degs(g[1], g[2])
head(deg_ui_ac$deg_results)
deg_ui_ac$plot
```

|               We can see many DEGs from the array, but the up- and down-regulations are not well contrasted. Using Bayesian moderation would yield a better separation between the two directions.<br>  
For each probe (i.e. first column in deg_result table),<br>

|               (1) logFC is the full changes between treatment times.  <br>
|               (2) CliffDelta is estimation on the effect size. <br> 
|               (3) W is the U-test statistic value. <br> 
|               (4) p.value is the p-value. <br> 
|               (5) adj.P.Value is the adjusted p-value for multiple testing across all probes.<br>

## Uninfected vs Chronic Controller
```{r perform deg Uninfected vs Chronic Controller, warning=FALSE}
g <- c("uninfected", "chronic_controller")
deg_ui_cc <- two_grp_degs(g[1], g[2])
head(deg_ui_cc$deg_results)
deg_ui_cc$plot
```

## Uninfected vs Progressor
```{r perform deg Uninfected vs Progressor, warning=FALSE}
g <- c("uninfected", "progressor")
deg_ui_pr <- two_grp_degs(g[1], g[2])
head(deg_ui_pr$deg_results)
deg_ui_pr$plot
```

## Asymtomatic Controller vs Chronic Controller
```{r perform deg Asymtomatic Controller vs Chronic Controller, warning=FALSE}
g <- c("asymptomatic_controller", "chronic_controller")
deg_ac_cc <- two_grp_degs(g[1], g[2])
head(deg_ac_cc$deg_results)
deg_ac_cc$plot
```

## Asymtomatic Controller vs Progressor
```{r perform deg Asymtomatic Controller vs Progressor, warning=FALSE}
g <- c("asymptomatic_controller", "progressor")
deg_ac_pr <- two_grp_degs(g[1], g[2])
head(deg_ac_pr$deg_results)
deg_ac_pr$plot
```

## Chronic Controller vs Progressor
```{r perform deg Chronic Controller vs Progressor, warning=FALSE}
g <- c("chronic_controller", "progressor")
deg_cc_pr <- two_grp_degs(g[1], g[2])
head(deg_cc_pr$deg_results)
deg_cc_pr$plot
```

# Enrichment/Overrepresentation Analysis
## Getting microarray annotation for gene ontology (GO)
|               The source article lists GPL32068 as the platform for annotation data, and it is described as being identical to the platform GPL16570. Although GPL32068 has a custom Chip Definition File (CDF), the annotations on GPL16570 are more comprehensive. Given that the two platforms contains the same annotation information, and I like the format of GPL16570, we will use GPL16570 as our source of GO annotations.  
```{r get GO data from GEO}
# download from GEO. 
# You may need to inspect the page to find the url
options(timeout = 600) # the file is about 240 mb
fp <- "gene_data/GPL16570-1802.txt"
url <- "https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?mode=raw&is_datatable=true&acc=GPL16570&id=1802&db=GeoDb_blob100"

if (!file.exists(fp)){
    download.file(url, fp)}

# load the text file as a table, and omit the metadata
GPL16570 <- read.table("gene_data/GPL16570-1802.txt", header = TRUE, sep = "\t",
                     skip = 0, quote = "", row.names = 1,fill = TRUE)
head(GPL16570)

# source study used lung tissues, and we shall filter the GO annotation related to lung data on GPL16570. Reasons are provided in the enrichment analysis section below
GPL16570_lung <- GPL16570 %>% 
  filter(grepl("lung", unigene, ignore.case = TRUE)) 
head(GPL16570_lung)

# note the number of relevant rows before and after filtering 
nrow(GPL16570)
nrow(GPL16570_lung)
options(timeout = 60) # reset the timeout parameter to default
```

## Create four sets of Gene Ontology (GO) terms 
```{r define parsing functions}
# gene id and titles are stored in the gene_assignment column in GPL16570_lung. 
# make a function to grab the second and third items from gene_assignment
parseGI <- function(gene_assignment){
  # use regex to find the first two items. I hate regex...
  str_ful <- str_match(gene_assignment, "// (\\w+) // (.+?) //")
  gene_id <- str_ful[2]
  gene_title <- str_ful[3]
  # return the data as list for subsequent cbind actions
  return(list(Gene_id = gene_id, Gene_title = gene_title))
}

# parsing GO terms
# biological process is chosen
parsGO <- function(probe, ontology = "GO_biological_process"){
  go_segment <- strsplit(GPL16570_lung[probe, "GO_biological_process"], "///")[[1]]
  go_id <- trimws(unlist(lapply(go_segment, function(x) strsplit(x, "//")[[1]][2])))
  go_description <- trimws(unlist(lapply(go_segment, function(x) strsplit(x, "//")[[1]][3])))
  go_source <- trimws(unlist(lapply(go_segment, function(x) strsplit(x, "//")[[1]][4])))

  # in case of probe without id (a.k.a. errors in data wrangling)
  if (length(go_id) > 0){
    go_tab <- cbind(Probe_id = probe,
                    Id = go_id,
                    Description = go_description,
                    Source = go_source)
    rownames(go_tab) <- rep(probe, nrow(go_tab))
    return(go_tab)
    return() # return nothing if go_id is empty/NA (perhaps stopping the process with a warning also works)
  }
}

# prepare AGO and GO data
# <NOTE this in inefficient. Try break down GO by Ex_group, and then define another function later to rbind them.>
prepGO <- function(degs){
  # match information between Array GO terms from DEG result and GPL16750_lung.
  AGO <- cbind(Gene_Symbol = GPL16570_lung[degs$Probe, c("probeset_id", "Gene_id", "Gene_title")],
               degs[, c("logFC", "p.value", "adj.P.Val")])
  colnames(AGO) <- c("Probe_id", "Gene_id", "Gene_title", "logFC", "P.Val", "adj.P.Val")
  
  # parse all GO terms on Biological Process column from GPL16570. 
  GO <- c()
  
  for (probe in rownames(AGO)){
    GO <- rbind(GO, parsGO(probe))}
  
  # drop the rows with only na.
  GO <- as.data.frame(GO[rowSums(is.na(GO)) != ncol(GO), ])
  GO <- na.omit(GO)
  
  # match AGO and GO, as DEGs without GO term is of little significant to our purpose.
  AGO <- AGO[which(AGO$Probe_id %in% GO$Probe_id), ]
  #AGO <- AGO %>% mutate(has_GO = ifelse(Probe_id %in% GO$Probe_id, TRUE, FALSE))
  
  # combine the result into a list for subsequent enrichment analysis.
  result <- list(AGO, GO)
  names(result) <- c("AGO", "GO")
  
  return(result)
}
```

```{r preping gene IDs and names from GPL16570}
# modify GPL16570_lung in place
GPL16570_lung <- GPL16570_lung %>%
  mutate(Gene_id = map_chr(gene_assignment, ~parseGI(.x)$Gene_id),
         Gene_title = map_chr(gene_assignment, ~parseGI(.x)$Gene_title))
```

|               This following section may be quite slow as we have many samples, probes, and GO terms. I should look into more efficient way to process the GO data in the future. As for now, please be patient with my insufficient brain.

<pre style="color: #7e9f7f; background-color: #222222; font-family: monospace; padding: 10px;">
                       _oo0oo_
                      o8888888o
                      88" . "88
                      (| -_- |)
                      0\  =  /0
                    ___/`---'\___
                  .' \\|     |# '.
                 / \\|||  :  |||# \
                / _||||| -:- |||||- \
               |   | \\\  -  #/ |   |
               | \_|  ''\---/''  |_/ |
               \  .-\__  '-'  ___/-. /
             ___'. .'  /--.--\  `. .'___
          ."" '<  `.___\_<|>_/___.' >' "".
         | | :  `- \`.;`\ _ /`;.`/ - ` : | |
         \  \ `_.   \_ __\ /__ _/   .-` /  /
     =====`-.____`.___ \_____/___.-`___.-'=====
                       `=---='

     ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
</pre>

```{r parsing gene and GO terms for Uninfected vs Asymtomatic Controller}
# specify the result from the correct group of deg.
degs <- deg_ui_ac$deg_results

# parse gene IDs and GO terms.
Genego_ui_ac <- prepGO(degs)
head(Genego_ui_ac$AGO)
head(Genego_ui_ac$GO)
```

```{r parsing gene and GO terms for Uninfected vs Chronic Controller}
# specify the result from the correct group of deg.
degs <- deg_ui_cc$deg_results

# parse gene IDs and GO terms.
Genego_ui_cc <- prepGO(degs)
head(Genego_ui_cc$AGO)
head(Genego_ui_cc$GO)
```

```{r parsing gene and GO terms for Uninfected vs Progressor}
# specify the result from the correct group of deg.
degs <- deg_ui_pr$deg_results

# parse gene IDs and GO terms.
Genego_ui_pr <- prepGO(degs)
head(Genego_ui_pr$AGO)
head(Genego_ui_pr$GO)
```

```{r parsing gene and GO terms for Asymtomatic Controller vs Chronic Controller}
# specify the result from the correct group of deg.
degs <- deg_ac_cc$deg_results

# parse gene IDs and GO terms.
Genego_ac_cc <- prepGO(degs)
head(Genego_ac_cc$AGO)
head(Genego_ac_cc$GO)
```

```{r parsing gene and GO terms for Asymtomatic Controller vs Progressor}
# specify the result from the correct group of deg.
degs <- deg_ac_pr$deg_results

# parse gene IDs and GO terms.
Genego_ac_pr <- prepGO(degs)
head(Genego_ac_pr$AGO)
head(Genego_ac_pr$GO)
```

```{r parsing gene and GO terms for Chronic Controller vs Progressor}
# specify the result from the correct group of deg.
degs <- deg_cc_pr$deg_results

# parse gene IDs and GO terms.
Genego_cc_pr <- prepGO(degs)
head(Genego_cc_pr$AGO)
head(Genego_cc_pr$GO)
```

## Fast analysis using genes and GO terms on org.Mm.eg.db (ClusterProfiler)
|               With a list of gene id and the logFC between two groups, we can perform an enrichment analysis based on org.Mm.eg.db. This block use Asymtomatic Controller and Progressor groups as an example. However, we should be cautious about two things when using this approach: (1) gene id from GEO may not have a match on org.Mm.eg.db; (2) org.Mm.eg.db contains all genes while the source article samples only lung tissues. Therefore, we may find signals of enrichment/overrepresentation not due to TB infections, but because we sample lung tissues!
```{r enrichment using clusterProfiler}
# filter the AGO list for DEGs with ah alpha level at 0.05, and grab the list of gene id.
gene_list <- Genego_ac_pr$AGO[which(Genego_ac_pr$AGO[, "adj.P.Val"] > 0), "Gene_id"]

# check the gene mapping with the database
# warning message at this step may due to deprecated genes, unannotated genes, or other issues.
mapped_genes <- bitr(gene_list, fromType = "SYMBOL", toType = "ENTREZID", OrgDb = org.Mm.eg.db)

# some genes are not found on org.Mm.eg.db, and we need to filter those out. They may be from more recent studies, and org.Mm.eg.db have mapped them.
unmapped_genes <- gene_list[!gene_list %in% mapped_genes$SYMBOL]
gene_list <- setdiff(gene_list, unmapped_genes)

tempt <- as.data.frame(gene_list)
tempt <- tempt %>%
  left_join(Genego_ac_pr$AGO, by = c("gene_list" = "Gene_id"))

gene_list <- tempt$logFC
names(gene_list) <- tempt$gene_list

# perform enrichment analysis.
orgdb_result <- enrichGO(gene = names(gene_list),
                         OrgDb = org.Mm.eg.db,
                         keyType = "SYMBOL",
                         ont = "BP",
                         pvalueCutoff  = 0.05, 
                         pAdjustMethod = "BH",
                         qvalueCutoff  = 0.05,
                         readable = TRUE)

# keep only the significant results by an alpha level of 0.05, and convert the result into a table.
orgdb_table <- as.data.frame(orgdb_result[orgdb_result$p.adjust < 0.05])
head(orgdb_table)
```

```{r clusterProfiler visualisation}
# clusterProfiler has many cool ways to visualise the enrichment analysis results, but the results generally would need to be in specific Class object like enrichResult.
# upsetplot
upset <- upsetplot(orgdb_result, n = 5) 
upset

# network plot
# get semData for the plot
sem_data <- godata('org.Mm.eg.db', ont = "BP", computeIC = TRUE)
p_ema <- emapplot(pairwise_termsim(orgdb_result, method = "Wang", semData = sem_data), showCategory = 10, color = "p.adjust", layout.params = list(layout = "nicely"), edge.params = list(min = 0), pie = "Count") 
p_ema
```

## Manual computation for enrichment analysis
### Theoretical Basis of Overrepresentation Analysis
|               The biological question of this report is whether and which gene ontology is overrepresented in the progressor group given the data from the microarray in the source study and the known gene annotations. This might lead to identification of meaningful and perhaps even clinically significant genes and/or biological processes.
|               The computation behind enrichment/overrepresentation analysis can be summarised by the following Venn diagram.
```{r Venn diagram of enrichment analysis theory, echo=FALSE, fig.width=6, fig.height=6}
# <NOTE first run of this chunk may cause error, run it again without modification solve the issue. Debugging required>
# <NOTE try 3-layer ggplot pie chart as you had trouble scaling the circles>
# Load necessary package
library(VennDiagram)
library(grid)

# define the sets with q and m as subsets of o
set_list <- list(
  q = c("q", "x"),        # q contains q and x
  m = c("m", "x"),        # m contains m and x
  o = c("q", "m", "x", "rest")  # o contains all elements
)

# generate a Venn diagram
vd <- venn.diagram(x = set_list,
                   category.names = c("q", "m", "o"),
                   filename = NULL,
                   lwd = 2,
                   col = "white",
                   fill = c("#102C57", "#B43F3F", "#E5D283"),
                   alpha = c(0.7, 0.8, 0.5),
                   cex = 0,
                   cat.cex = 2,
                   cat.pos = c(-10, -15, 18),
                   cat.dist = c(0.02, 0.03, -0.05),
                   cat.fontface = "bold",
                   cat.default.pos = "text")

# show the diagram
grid.draw(vd)

# add a label for the intersection between m and q
labelx <- textGrob("x", 
                   x = unit(0.5, "npc"), 
                   y = unit(0.40, "npc"),
                   gp = gpar(fontsize = 18, 
                             fontface = "bold", 
                             col = "white"))
grid.draw(labelx)
```

Notation of the Venn diagram:<br>

|               $\textbf{o}$ = number of probes from the microarray<br>
|               $\textbf{m}$ = number of probes from the microarray with known gene ontology (background count)<br>
|               $\textbf{q}$ = number of probes in our list of differentially expressed gene (foreground count)<br>
|               $\textbf{x}$ ($q \cap m$) = number of differentially expressed gene with the selected gene ontology<br>
|               $\textbf{n}$ = number of probes not in our lit of differentially expressed gene<br>
|               $\text{Foreground ratio}$ = $\frac{x}{q}$<br>
|               $\text{Background ratio}$ = $\frac{m}{o}$<br>

|               Our biological question above is now transformed into a comparison of whether the $\underline{\text{Foreground ratio}}$ is $\textbf{significantly}$ larger than the $\underline{\text{Background ratio}}$.

|               Microarray data is huge, especially with 134 samples. Counting and comuting the ratio manually would most certainly be impractical. Instead, we can apply a form of the hypergeometric test/distribution in R.

|               A story motivation of the hypergeometric distribution can be given as the following: A box contains i pieces of chocolates and j pieces of pickles. Let's consider getting chocolates are successes and getting pickles are failures. Assuming any object from the box is equally likely go be drawn, the random variable X refers to the number of successes from a draw of n objects from the box without replacement. 

|               In the context of hypergeometric distribution, the question of enrichment analysis is again transformed into the following scenario: We select $\textbf{q}$ probes from the entire microarray $\textbf{o}$, the probability ($\textbf{P}$) of finding $\textbf{x}$ or more probes with differentially expressed gene and gene ontology is given as:
$$
P = \sum_{i=x}^{min(q,m)} \binom{q}{x}\binom{o-k}{m-x}
$$

|               Probability Mass Function from other sources may have an additional denominator, namely $\binom{q+o-k}{m}$. This denominator is a scaling term, to ensure that the resulting probability is between 0 and 1. In our enrichment analysis, the goal is compare the relative probabilities among the differential expressed genes. The scaling term can be omitted for such purpose. That said, some packages may implement the algorithm in a more rigorous way, and we could see the scaling denominator in some practices.
|               That may be enough theory now. Let's get into some R-actions.

## Define a Function for Computation
|               At this point, we have both the AGO from the DEG analysis among the various pairs of contrast groups and the GO data from platform on GEO. Instead of using clusterProfiler, we can manually perform the enrichment/overrepresentation analysis. The theoretical section at the beginning requires computation of the background count $\textbf{m}$, the foreground count $\textbf{q}$, the DEGs of selected GO term $\textbf{x}$, and the number of probes without GO terms $\textbf{n}$. An alpha level of 0.05 is chosen. A different alpha level can be chosen based on the tolerance for Type I and II errors.
```{r define an EA function to perform the computation}
EACompute <- function(ago, go){
  # assign the arguments.
  AGO <-  ago
  GOdata <- go
  
  # check the minimal adj.P.Val.
  if (min(AGO$adj.P.Val) < 0.05){
    # perform the EA computation.
    # compute for foreground.
    fg_probe <- rownames(AGO)[which(AGO[, "adj.P.Val"] < 0.05)]
    fg_go_id <- GOdata$Probe_id[GOdata$Probe_id %in% fg_probe]
    
    # compute for background.
    bg_probe <- rownames(AGO)
    bg_go_id <- GOdata$Probe_id[GOdata$Probe_id %in% bg_probe]
    
    # compute for the counts for relevant terms form the theoretical section for each gene ontology.
    EA <- data.frame() # enrichment analysis
    
    # counting for each GO term.
    for (Id in unique(fg_go_id)){
      # total count in background with GO term
      m <- sum(bg_go_id == Id) 
      # total count in background without GO term
      n <- length(bg_go_id) - m 
      # total count in foreground
      q <- length(fg_go_id)
      # total count in foreground with GO term
      x <- sum(fg_go_id == Id)
      # compute the hypergeometric p-value
      p_val <- phyper(x - 1, m, n, q, lower.tail = FALSE)
      # combine the result for all probes
      EA <- rbind(EA, c(m, m + n, q, x, p_val))
    }
    rownames(EA) = unique(fg_go_id)
    colnames(EA) = c("m", "o(m + n)", "q", "x", "P.Val.hygeo")
    
    if(nrow(EA) > 0 && any(!is.na(EA[, "x"]))){
      # adjust the P.Val.hygeo for multiple testing across the entire array of probes. Finding true positive is more important to me in this analysis, but use one of the other adjusting methods as needed.
      EA <- cbind(EA, "adj.P.Val.hygeo" = p.adjust(EA[, "P.Val.hygeo"], "BH"))
      
      # filter again for only the significant results by adj.P.Val.hygeo. We use 0.05 as the alpha level as 0.05 captures only one GO term...
      GOsig <- rownames(EA)[which(EA[, "adj.P.Val.hygeo"] < 0.05)]
      
      # grab and add GO data.
      GOname <- data.frame(Probe_id = character(), Description = character(), stringsAsFactors = FALSE)

      for (go in GOsig){
        description <- unique(GOdata[which(GOdata[, "Probe_id"] == go), "Description"])
        df <- data.frame(Probe_id = go, Description = description)
        GOname <- rbind(GOname, df)}
      
      EA <- EA %>% 
        rownames_to_column(var = "Probe_id") %>%
        left_join(GOname, by = "Probe_id")
      
      # we can format the enrichment analysis output as observed and expected, which may be easier to interpret.
      # recall that we want to compare the foreground and background ratios. DEG with the selected gene ontology/foreground count.
      # foreground ratio.
      Ratio_fg <- as.numeric(EA[, "x"])/as.numeric(EA[, "q"])
      
      # background ratio.
      Ratio_bg <- as.numeric(EA[, "m"])/as.numeric(EA[, "o(m + n)"])
      
      # expected value of foreground based on the background ratio.
      Ratio_mu_bg <- round(Ratio_bg * as.numeric(EA[, "q"]), 5)
      
      # create a dataframe as a readable result.
      EA <- data.frame(cbind(`GO id` = EA[, "Probe_id"],
                             Description = EA[, "Description"],
                             `Expected Frequency` = paste0(Ratio_mu_bg, "/", EA[, "q"]),
                             `Observed Frequency` = paste0(EA[, "m"], "/", EA[, "q"]),
                             `Background Ratio` = paste0(EA[, "m"], "/", EA[, "o(m + n)"]),
                             `FoldChange` = Ratio_fg/Ratio_bg, # of GO term.
                             `P.Val` = as.numeric(EA[, "P.Val.hygeo"]),
                             `P.Val.adjusted` = as.numeric(EA[, "adj.P.Val.hygeo"])))
      
      # add gene information
      tempt <- EA %>%
        left_join(as.data.frame(cbind(probe_id = rownames(GOdata), GOdata)), by = c("GO.id" = "Id"))
      
      AGO$Probe_id <- as.character(AGO$Probe_id)
      
      EA_join <- tempt %>%
        left_join(AGO, by = c("GO.id" = "Probe_id"))
      
      # remove duplicates and unnecessary columns
      EA_join <- select(EA_join, c("probe_id", "GO.id", "Gene_id",  
                                   "logFC", "Description.x", "FoldChange", 
                                   "P.Val.x", "P.Val.adjusted", "Observed.Frequency", "Background.Ratio"))
      
      colnames(EA_join) <- c("probe_id", "GO.id", "Gene_id", 
                             "logFC", "Description", "FoldChange", 
                             "P.Val", "P.Val.adjusted", "Observed.Frequency", "Background.Ratio")
      
      EA_join <- unique(EA_join)
      
      # remove duplicate for a clean output table.
      tempt <- EA %>%
        select(c("GO.id", "Description", "Expected.Frequency", "Observed.Frequency", 
                 "Background.Ratio", "FoldChange", "P.Val", "P.Val.adjusted"))
      
      join_list <- EA_join %>%
        select(c("GO.id", "Gene_id"))
      
      tempt <- tempt %>%
        left_join(join_list, by = "GO.id") 
      
      tempt <- tempt %>%
        group_by(GO.id) %>%
        dplyr::reframe(
          Gene.id = paste(Gene_id, collapse = "/"),
          Description = paste(Description, collapse = "/")) 
      
      # drop the Desirption from EA first (this is inefficient)
      EA <- EA %>% select(-Description)
      EA_tab <- tempt %>%
        left_join(EA, by = "GO.id")
      
      # remove duplicate gene id and biological pathways
      EA_tab <- EA_tab %>%
        mutate(Gene.id = sapply(Gene.id, function(x) paste(unique(str_split(x, "/")[[1]]), collapse = "/"))) %>% 
        mutate(Description = sapply(Description, function(x) paste(unique(str_split(x, "/")[[1]]), collapse = "/"))) %>%
        distinct()

      # create an enrichResult object if you want to use some packages for visualisation.
      # create dataframe for encichResult obj.
      enrich_df <- select(EA_join, c("probe_id", "GO.id", "Description", "Gene_id", "logFC", "P.Val", "P.Val.adjusted"))
      tempt <- GPL16570_lung[, c("probeset_id", "Gene_id")]
      tempt$probe_id <- as.character(rownames(tempt))
      
      enrich_df <- enrich_df %>%
        left_join(tempt, by = c("GO.id" = "probe_id")) %>%
        select(c("GO.id", "Description", "Gene_id.x", "Gene_id.y", "logFC", "P.Val", "P.Val.adjusted")) %>% 
        na.omit()
      
      # set the column names to avoid confusions.
      colnames(enrich_df) <- c("GO.id", "Description", "AGene.id", "BgGene.id", "logFC", "P.Val", "P.Val.adjusted")
      
      # count the number of genes and concat the gene id column. The numerator of foreground and background gene ratios are supposed to be the same as we filtered probes by GO terms in earlier sections.
      enrich_df <- enrich_df %>%
        group_by(GO.id, Description, P.Val, P.Val.adjusted) %>%
        dplyr::summarize(
          AGene.id = paste(AGene.id, collapse = "/"),
          AGene.count = sapply(strsplit(AGene.id, "/"), length),
          BgGene.id = paste(BgGene.id, collapse = "/"),
          BgGene.count = sapply(strsplit(BgGene.id, "/"), length))
      
      array_gene_list <- unique(EA_join$Gene_id)
      gpl_gene_list <- unique(GPL16570_lung$Gene_id)
      
      # get denominator for computation of gene ratio.
      array_gene_count <- length(unique(EA_join$Gene_id))
      gpl_gene_count <- length(unique(GPL16570_lung$Gene_id))
      
      # wrangle data into a dataframe with correct format.
      ea_df <- data.frame(cbind(
        ID = enrich_df[, "GO.id"],
        Description = enrich_df[, "Description"],
        GeneRatio = paste0(enrich_df$AGene.count, "/", array_gene_count),
        BgRatio = paste0(enrich_df$BgGene.count, "/", gpl_gene_count),
        pvalue = as.numeric(unlist(enrich_df[, "P.Val"])),
        p.adjust = as.numeric(unlist(enrich_df[, "P.Val.adjusted"])),
        qvalue = rep(0.03, nrow(enrich_df)), # a dummy value for plotting, p.adjust are too small for calculating qvalue.
        geneID = enrich_df[, "AGene.id"],
        Count = enrich_df[, "AGene.count"],
        stringsAsFactors = FALSE))
      
      # ensure the column ID are exact.
      colnames(ea_df) <- c("ID", "Description", "GeneRatio", "BgRatio", "pvalue", "p.adjust", "qvalue", "geneID", "Count")
      
      # create a geneset list for mapping GO terms and gene ID.
      geneSets <- setNames(lapply(ea_df$geneID, function(x) unlist(strsplit(x, "/"))), ea_df$ID)
      
      # create enrichResult object.
      EA_obj <- new("enrichResult",
                    result = ea_df,
                    pvalueCutoff = 0.05,
                    pAdjustMethod = "BH",
                    qvalueCutoff = 0.02,
                    gene = array_gene_list,
                    universe = character(0),
                    gene2Symbol = character(0),
                    geneSets = geneSets,
                    organism = "Mus musculus",
                    keytype = "SYMBOL",
                    ontology = "BP", # biological process
                    readable = TRUE)
      
      result <- list(EA_tab, EA_obj)
      names(result) <- c("EA_tab", "EA_obj")
      return(result)
    }else print("Not enougn significant GO terms for computation. Increased sample size is required.")
  }else print("GO terms from contrast groups are not significant enough for computation. Set a larger alpha level to proceed.")
}
```

## Perform Pairwise Enrichment/Overrepresentation Analysis
```{r Overrepresentation Uninfected vs Asymtomatic Controller}
EA_ui_ac <- EACompute(Genego_ui_ac$AGO, Genego_ui_ac$GO)
head(EA_ui_ac$EA_tab)

# save the result as a csv file if needed.
# write.csv(as.data.frame(EA_ui_ac$EA_tab), "EA_Fin_ui_ac.csv", row.names = FALSE)
```

```{r Overrepresentation Uninfected vs Chronic Controller}
EA_ui_cc <- EACompute(Genego_ui_cc$AGO, Genego_ui_cc$GO)
head(EA_ui_cc$EA_tab)
```

```{r Overrepresentation Uninfected vs Progressor}
EA_ui_pr <- EACompute(Genego_ui_pr$AGO, Genego_ui_pr$GO)
head(EA_ui_pr$EA_tab)
```

```{r Overrepresentation Asymtomatic Controller vs Chronic Controller}
EA_ac_cc <- EACompute(Genego_ac_cc$AGO, Genego_ac_cc$GO)
head(EA_ac_cc$EA_tab)
```

```{r Overrepresentation Asymtomatic Controller vs Progressor}
EA_ac_pr <- EACompute(Genego_ac_pr$AGO, Genego_ac_pr$GO)
head(EA_ac_pr$EA_tab)
```

```{r Overrepresentation Chronic Controller vs Progressor}
EA_cc_pr <- EACompute(Genego_cc_pr$AGO, Genego_cc_pr$GO)
head(EA_cc_pr$EA_tab)
```
Voil! Now we have the differentially expressed probes with their gene ontology for biological process from our microarray!<br>

## If Genes Reported in Sources Study Present? 
|               Source study reports on comparison between asymptomatic and other groups, as one of the key finding is that B-cell pathways genes and CD20+ genes are overrepresented in the asymptomatic group. As such, this section will only compare asymtopmatic and any given group.
```{r Uninfected vs Asymtomatic Controller}
com_gene_tab_ui_ac <- EA_ui_ac$EA_tab %>%
  filter(str_to_lower(Gene.id) %in% str_to_lower(c("CD79B", "CD79A", "CR2", "FLT3", "BANK1", "PAX5", "MS4A1", "CD19", "CD22", "NUP107", "RAE1")))


com_biopath_tab_ui_ac <- EA_ui_ac$EA_tab %>%
  filter(str_detect(Description, regex("B cell activation|B cell differentiation|Lymphocyte differentiation|B cell proliferation|Lymphocyte proliferation|B cell receptor signaling pathway|Regulation of B cell activation|Transcription-dependent tethering of RNA polymerase II gene DNA at nuclear periphery", ignore_case = TRUE)))

com_gene_tab_ui_ac
com_biopath_tab_ui_ac
```

```{r Asymtomatic Controller vs Chronic Controller}
com_gene_tab_ac_cc <- EA_ac_cc$EA_tab %>%
  filter(str_to_lower(Gene.id) %in% str_to_lower(c("CD79B", "CD79A", "CR2", "FLT3", "BANK1", "PAX5", "MS4A1", "CD19", "CD22", "NUP107", "RAE1")))

com_biopath_tab_ac_cc <- EA_ac_cc$EA_tab %>%
  filter(str_detect(Description, regex("B cell activation|B cell differentiation|Lymphocyte differentiation|B cell proliferation|Lymphocyte proliferation|B cell receptor signaling pathway|Regulation of B cell activation|Transcription-dependent tethering of RNA polymerase II gene DNA at nuclear periphery", ignore_case = TRUE)))

com_gene_tab_ac_cc
com_biopath_tab_ac_cc
```

```{r Asymtomatic Controller vs Progressor}
com_gene_tab_ac_pr <- EA_ac_pr$EA_tab %>%
  filter(str_to_lower(Gene.id) %in% str_to_lower(c("CD79B", "CD79A", "CR2", "FLT3", "BANK1", "PAX5", "MS4A1", "CD19", "CD22", "NUP107", "RAE1")))

com_biopath_tab_ac_pr <- EA_ac_pr$EA_tab %>%
  filter(str_detect(Description, regex("B cell activation|B cell differentiation|Lymphocyte differentiation|B cell proliferation|Lymphocyte proliferation|B cell receptor signaling pathway|Regulation of B cell activation|Transcription-dependent tethering of RNA polymerase II gene DNA at nuclear periphery", ignore_case = TRUE)))

com_gene_tab_ac_pr
com_biopath_tab_ac_pr
```

# Conclusions
|               (1) Source data is non-normal with some outliers, which justifies the use of Mann-Whitney U-test.<br>
|               (2) Of the 11 overrepresented genes reported in the asymptomatic group in the source article, 4 are recovered from this project, namely RAE1, BANK1, NUP107, and CD79B.<br>
|               (3) Of the 11 overrepresented genes reported in the source article, asymptomatic controller do not show any of the 11 genes when compared to the chronic controller group. <br>
|               (4) Many pathways are found to be overrepresented between asymptomatic controller show 249 genes with a wide range of pathways, instead of strict B-cell pathways. <br>

# Final thoughts
|               (1) The raw CEL files show a high level of correlation even before normalisation, which may suggest the variability between groups is somewhat low. By using rma() in this exploration, source data might have been over,normalisaion, causing the FoldChange computation in the overrepresentation (manual computation) analysis to be identical across all probes in a given pairwise comparison.I attempted downsampling but the issues persist. A better technique with higher sensitivity may be necessary for this source data.<br>
|               (2) Some overrepresented genes (e.g., Bank1, Rae1) lack annotations on GPL16570, which might be one of the reasons why the source article uses a custom annotation platform.<br>